Type-Directed, Whitespace-Delimited Parsing for Embedded DSLs

### 2013-06-14 globaldsl13/motivation.tex
GlobalDSL13 final

\section{Motivating Examples}
\label{s:motivation}

\begin{figure}
  \centering
  \begin{lstlisting}
val hotProductDashboard = new Application
    external component twitter : Feed
        location [www.twitter.com]
    external component client : Browser
        connects to servlet
    component servlet : DashServlet
        connects to productDB, twitter
        location [intranet.nameless.com]
    component productDB : Database
        location [db.nameless.com]
    policy NamelessSecurity
        must salt [servlet.login.password]
        connect * -> servlet with HTTPS
        connect servlet -> productDB with TLS
  \end{lstlisting}
  \caption{Wyvern DSL: Architecture Specification}
  \label{f:dsl-arch}
\end{figure}

We start with a few examples meant to illustrate the expressiveness of our approach and the breadth of DSLs we plan for it to support.  The examples are presented in the proposed syntax for Wyvern, a new language being developed by our group  initially targeted toward building secure web and mobile applications. We will further clarify how these examples are parsed in Section \ref{s:approach}.

The first example, shown in Figure~\ref{f:dsl-arch}, describes the overall architecture of a ``hot product dashboard'' application.  The constructor \verb|Application| takes an argument of type \verb|ComponentArch|. Rather than being provided explicitly, a DSL for specifying the component architecture of the application is used to provide this argument because it appears in a whitespace-delimited block. The example architecture declares several components, some of which are marked \keyw{external} to indicate that they are used by this application but are not part of it directly. Component types, connectivity and properties, such as location, are declared within further-indented blocks.

The \keyw{policy} keyword introduces a security policy, which constrains the communication protocols to be used as well as
enforces secure handling of passwords.  Again, indentation delimits
the scope of the security policy.  This
%example
illustrates that one DSL may be nested within another.

\begin{figure}
  \centering
  \begin{lstlisting}
val newProds = productDB.query 
    select twHandle 
    where introduced - today < 3 months
this.prodTwt = Feed(newProds)
return prodTwt.query
    select *
    group by followed
    where count > POPULAR  
  \end{lstlisting}
  \caption{Wyvern DSL: Queries}
  \label{f:dsl-query}
\end{figure}

Figure~\ref{f:dsl-query} shows how a DSL for querying can be used from within ordinary Wyvern code.  The example shows the body of a method for computing a feed that is derived from tweets about the company's new products.  In this example, the use of a querying DSL is triggered by the use of a function, query, expecting a \verb|DBQuery|.  This type defines a particular syntax for database queries, delimited by the indented block as the reader would expect. This is similar to what can be expressed with built-in query syntax such as Microsoft's LINQ \cite{mslinq}, but in this case, defined in a user-defined type. The feed query operator may define different features, as seen on Line 5.
\begin{figure}
  \centering
  \begin{lstlisting}
html:
    head:
        title: [text "Hot Products"]
        style:
            myStylesheet
    body:
        div id="search":
            SearchBox("products")
        div id="products":
            FeedBox(servlet.hotProds())
  \end{lstlisting}
  \caption{Wyvern DSL: Presentation}
  \label{f:dsl-presentation}
\end{figure}

Finally, Figure~\ref{f:dsl-presentation} shows a DSL for rendering the hot product application in a web browser.  The DSL is based on HTML but uses our indentation-based syntax and allows Wyvern code to be integrated -- function/method calls and variables of the appropriate type can be used at any point. Literal tags are introduced by a tag name followed by any tag attributes followed by a colon.

>>>>

% !TEX root = globaldsl13.tex
\section{Motivating Examples}
\label{s:motivation}

\begin{figure}
  \centering
  \begin{lstlisting}
val dashboardArchitecture : Architecture = ~
    external component twitter : Feed
        location www.twitter.com
    external component client : Browser
        connects to servlet
    component servlet : DashServlet
        connects to productDB, twitter
        location intranet.nameless.com
    component productDB : Database
        location db.nameless.com
    policy mainPolicy = ~
        must salt servlet.login.password
        connect * -> servlet with HTTPS
        connect servlet -> productDB with TLS
  \end{lstlisting}
  \caption{Wyvern DSL: Architecture Specification}
  \label{f:dsl-arch}
\end{figure}

We start with a few examples to illustrate the expressiveness of our approach and the breadth of DSLs we plan for it to support.  The examples are presented in the proposed syntax for Wyvern, a new language being developed by our group that is targeted toward building secure web and mobile applications. We will informally describe each of these examples here, and further explain how such code is parsed in Section \ref{s:approach}.

The first example, shown in Figure~\ref{f:dsl-arch}, describes the overall architecture of a ``hot product dashboard'' application.  The variable \lstinline{dashboardArchitecture} is explicitly ascribed type \lstinline{Architecture}. Rather than explicitly providing a value of this type, we instead use a DSL that makes specifying the component architecture of the application more concise and readable. This DSL code appears in the subsequent whitespace-delimited block and is introduced by a tilde (\lstinline{~}). The example architecture declares several components, some of which are declared \keyw{external} to indicate that they are used by this application but are not part of it directly. Component types are declared after a colon and attributes like connectivity location, are declared after the type (formatted in an indented block for readability). 
The \keyw{policy} keyword (line 11) introduces a security policy, which constrains the communication protocols that can be used and 
enforces the secure handling of passwords. A separate type, \lstinline{Policy}, is associated with such policies. Although we could instantiate this type explicitly using a Wyvern expression, we use a DSL for defining policies instead, again within a whitespace-delimited block introduced by a tilde.

\begin{figure}
  \centering
  \begin{lstlisting}
val newProds = productDB.query(~)
    select twHandle 
    where introduced - today < 3 months
val prodTwt = new Feed(newProds)
return prodTwt.query(~)
    select *
    group by followed
    where count > 1000
  \end{lstlisting}
  \caption{Wyvern DSL: Queries}
  \label{f:dsl-query}
\end{figure}

Figure~\ref{f:dsl-query} shows how a DSL for database queries can be used from within ordinary Wyvern code.  The example shows code for computing a feed that is derived from tweets about a company's new products.  In this example, the use of a querying DSL is triggered by the use of methods named \lstinline{query} expecting an argument of type \lstinline{DBQuery} (line 1) or \lstinline{FeedQuery} (line 5) respectively.  These types define related but distinct syntax for queries, determined by the expected type of expression where the tilde appears (tildes need not appear only at the ends of lines). Queries are again delimited by indentation. This mechanism is similar to what can be expressed in languages with built-in query syntax like LINQ \cite{mslinq}, but in this case, it is entirely user-defined, rather than built into the language.\begin{figure}
  \centering
  \begin{lstlisting}
serve(page, loc) where 
  val page = ~ 
    html:
      head:
        title: Hot Products
        style: {myStylesheet}
        body:
          div id="search":
            {SearchBox("products")}
          div id="products":
            {FeedBox(servlet.hotProds())}
  val loc = ~
    products.nameless.com
  \end{lstlisting}
  \caption{Wyvern DSLs: Presentation and URLs}
  \label{f:dsl-presentation}
\end{figure}

Finally, Figure~\ref{f:dsl-presentation} shows a DSL for presenting the hot product application to a web browser, served at a particular URL. Here, two DSLs are used within a single function call. To allow this without introducing ambiguity, the user can use a \keyw{where} clause, similar to that found in Haskell \cite{jones2003haskell}. The presentation DSL is based on HTML and associated with a type, \lstinline{HTMLElement}. It uses an indentation-sensitive syntax and allows integration of Wyvern code of the appropriate type using curly braces. The second DSL simply canonicalizes URL literals into Wyvern values of type \lstinline{URL}.

### 2013-06-14 globaldsl13/related.tex
GlobalDSL13 final

\section{Related Work}
\label{s:related}
The most common mechanism for creating programming language extensions is macros, as exemplified by hygienic macros in Scheme. Macros in Scheme and other Lisp-style languages are written in Lisp itself, and benefit from the regular syntax of the underlying language and the consistent use of parentheses as expression delimiters.  The syntactic nature of macros allows them to control parsing within a delimited scope, and this delimited control of parsing inspired our approach in Wyvern.  Wyvern's use of types to trigger custom parsers avoids the Lisp limitation of starting each DSL with a keyword or macro name.

Wyvern chooses whitespace delimitation rather than parenthesis delimitation because, anecdotally, many programmers find syntax in that style to be more readable.  There are several proposals for making Lisp and its dialects whitespace-delimited, including sweet-expressions \cite{sweet-expr} and SRFI-49 \cite{srfi-49} for Scheme, P4P proposal for Racket \cite{p4p-proposal}, Kernel's f-expressions \cite{fexprthesis} for Common Lisp \cite{f-expr}, and others. None of them seem to be successfully implemented and deployed within the Lisp communities, however.  This may be because Lisp programmers don't want to switch away from parentheses, yet having an alternate syntax is not by itself enough to draw non-Lisp programmers into the community.

@@

Some language extensibility projects provide extension at levels of abstraction above parsing. For instance, OJ (previously, OpenJava)
\cite{Tatsubori00openjava:a} provides a macro system based on a meta-object
protocol, and Backstage Java~\cite{Palmer:2011:BJM:2048066.2048137} employs compile-time meta-programming.  Both of these systems provide macro-style rewriting of existing source code, but they provide at most limited extension of language parsing.

Other systems aim at providing more flexible forms of syntax extension compared to our whitespace-delimited approach, at the potential cost of complexity or conflicts.  Camlp4 \cite{camlp4} is a preprocessor for OCaml that offers the developer the ability to extend the concrete syntax of the language via the use of parsers and extensible grammars.  SugarJ \cite{Erdweg:2011:SLL:2048147.2048199} takes a library-centric approach which allows to syntactically extend the Java language by adding libraries. In Wyvern, the core language (that is, the \verb|Exp| sort) does not get extended directly, so conflicts cannot arise and reasoning about which DSL is applicable is much simpler.

Finally, researchers have developed a number of DSL frameworks and workbenches, such as the Meta Programming System (MPS) \cite{mps}, Spoofax \cite{KatsVisser2010}, the Intentional Domain Workbench (IDW) \cite{idw}, and Ens\={o} \cite{enso}, that provide support for extending a variety of programming languages by applying generic rules and concepts.  The Marco language \cite{lee:2012:marco} similarly provides macro definition at a level of abstraction that is largely independent of the target language.

Compared to these approaches, Wyvern focuses on making one programming language extensible with embedded DSLs, provides a common lexer and extensible parsing confined to whitespace-delimited regions, and investigates a novel type-based approach to triggering parser extensions.

>>>>

% !TEX root = globaldsl13.tex
\section{Related Work}
\label{s:related}
The most well-known mechanism for extending languages is macros, as exemplified by hygienic macros in Scheme. Macros in Scheme and other Lisp-style languages are written in the language itself and benefit from its simple syntax -- parentheses universally serve as expression delimiters (although proposals for whitespace as a substitute for parentheses have been made \cite{srfi-49}). Our work is inspired by this flexibility, but aims to support richer syntax as well as static types. Wyvern's use of types to trigger parsing  avoids the overhead of needing to invoke macros explicitly by name and makes it easier to compose DSLs declaratively.

@@

Some language extensibility projects provide metaprogramming facilities at levels of abstraction above parsing. For instance, OJ (previously, OpenJava)
\cite{Tatsubori00openjava:a} provides a macro system based on a meta-object
protocol, and Backstage Java~\cite{Palmer:2011:BJM:2048066.2048137}, Template Haskell \cite{sheard2002template} and others employ compile-time meta-programming.  Each of these systems provide macro-style rewriting of source code, but they provide at most limited extension of language parsing.

Other systems aim at providing forms of syntax extension that change the base language, as opposed to our whitespace-delimited approach.  For example, Camlp4 \cite{camlp4} is a preprocessor for OCaml that offers the developer the ability to extend the concrete syntax of the language via the use of parsers and extensible grammars.  SugarJ \cite{Erdweg:2011:SLL:2048147.2048199} takes a library-centric approach which supports syntactic extension of the Java language by adding libraries. In Wyvern, the core language (particularly the \keyw{EXP} sort) is not extended directly, so conflicts cannot arise at link-time.

Scoping DSLs to expressions of a single type comes at the expense of some flexibility, but we believe that many uses of DSLs are of this form already. A previous approach has considered type-based disambiguation of parse forests for supporting quotation and anti-quotation of arbitrary object languages \cite{bravenboer2005generalized}. Our work is similar in spirit, but does not rely on generation of parse forests and associates grammars with types, rather than types with grammar productions. We believe that this is a more simple and flexible methodology. The remaining approaches to syntax extension, such as XJ \cite{DBLP:conf/scam/ClarkSW08} are keyword-directed in some form. We believe that a type-directed approach is more seamless and general, sacrificing a small amount of identifiability in some cases.

Researchers have also developed DSL frameworks and language workbenches, including MPS \cite{mps}, Spoofax \cite{KatsVisser2010}, Ens\={o} \cite{enso} and others \cite{krahn2008monticore,van1992pregmatic} that provide support for generating new programming languages and tooling in a modular manner.  
%The Marco language \cite{lee:2012:marco} similarly provides macro definition at a level of abstraction that is largely independent of the target language.
Compared to these approaches, Wyvern focuses on extensibility \emph{internal} to the language, rather than taking an approach where each DSL is \emph{external} relative to the host language, improving interoperability and composability.

Finally, recent work on Active Code Completion is related to this work in that it associates code completion palettes with types \cite{omar2012active}. Such palettes could be used for defining a DSL syntax for types. However, that syntax is  immediately translated to Java syntax at edit-time while this work integrates with the core parsing facilities of the language.
